Type theory in type theory with single substitutions

### 2025-01-30 cpp/p.tex
stuff

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck the other equations. SSC provides a simple, minimalistic
alternative to parallel substitution calculi or B-systems for defining
type theory. SSC relates to CwF as extensional combinatory calculus
relates to lambda calculus: there are more models of the former, but
the syntaxes are equivalent. If we have some additional type formers,
we show that SSC models are actually equivalent to CwF-based models in
a weak sense. Most results in this paper were formalised in Agda.
\end{abstract}

@@

Some advantages of using a GAT presentation:
\begin{itemize}
\item  In type theory, typing depends on conversion:
  in an extrinsic presentation, this is expressed by a rule such as
  the following.
  \[
  \infer{\Gamma\vdash t : B}{\Gamma\vdash A \sim B && \Gamma\vdash t : A}
  \]
  When conversion $\sim$ is equality, this rule holds by the general
  properties of equality; and we also avoid lots of boilerplate
  stating that e.g.\ conversion is a congruence with respect to all
  operations (for an intrinsic presentation with explicit conversion
  relation see \cite{chapman09eatitself}).
\item Syntax and semantics are concordant: the syntax is simply the
  initial model, which always exists for any GAT and is called a
  quotient inductive-inductive type
  \cite{DBLP:journals/pacmpl/KaposiKA19}. The fact that the syntax is
  quotiented means that any function defined on it needs to respect
  the equations. For example, it is not possible to define a printing
  function which distinguishes the terms $(\lambda x.x)\,1$ and $1$,
  as these are convertible. However, it is still possible to define
  normalisation \cite{lmcs:4005} and typechecking \cite{typecheck} at
  this level of abstraction.
\item The GAT approach is also more abstract than extrinsic
  approaches, which means that we don't have to make ad-hoc
  choices. For example, we do not have to decide whether we want
  function space {\`a} la Curry or {\`a} la Church; whether we work
  with paranoid or economic typing rules \cite[Section
    5]{winterhalter:hal-01626651}; whether we have explicit or
  recursively defined substitution (instantiation of variables by
  terms, written $t[x\mapsto u]$ or $t[u/x]$).  These choices are
  forced by the notion of GAT: the domain of a lambda appears in its
  type, so we are always {\`a} la Church; algebraic theories don't
  allow recursively defined operators, so instantiation is an explicit
  operation; and so on.
\end{itemize}

@@

  \node (c2) at (5,1) {$\tilde{B}_1$};

@@

Using our notation, a B-frame contains the following data. \\

@@

Kaposi and Luksa \cite{luksa} defined a telescopic single substitution
calculus for simple type theory, it can be seen as a simply typed
version of B-systems. They show that the category of contextual models
of their calculus is equivalent to the category of contextual simply
typed CwFs with function space. Their equivalence result holds even
without the presence of function space. In the presence of function
space, the same equivalence holds for the simply typed version of our
substitution calculus, see the formalisation accompanying this paper.

@@

\paragraph{Formalisations of type theory}
Most computer formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than plain Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. Although we don't have a general
proof, we have experimental evidence that QIITs are supported by the
cubical set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid
model \cite{kaposi-qiit-setoid}, thus Cubical Agda
\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
theory have been formalised \cite{cohtt}. Without QIITs, tricks such
as shallow embedding the syntax can be used to formalise metatheoretic
results about type theory such as gluing
\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

\paragraph{SOGATs}
We can define languages with binders as second-order theories
following higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99}
and logical frameworks
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}. Second-order
generalised algebraic theories (SOGATs
\cite{uemura,DBLP:conf/fscd/KaposiX24}) can be used to present type
theory more abstractly than GATs: SOGATs abstract over the exact
definition of the substitution calculus (whether substitutions are
single or parallel; whether we have CwF-style or Ehrhard-style
operations): contexts, substitutions, instantiations are simply
modelled by the metatheoretic function space. For example, a type
theory with $\Pi$, Coquand-universes and a universe level lifting
operation is described by the following SOGAT (where $\cong$ means
isomorphism):
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
% Adding $\top$ and $\Sigma$ types (Definition \ref{def:sigma}) is
% described as the following extension of the above SOGAT:
% \begin{equation}\label{eq:sigma}
%   \begin{alignedat}{10}
%     & \top && : \Ty\,0 \\
%     & \tt && : \Tm\,\top \\
%     & \top\eta && : t = \tt \\
%     & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
%     & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
% \end{alignedat}
% \end{equation}
Second-order models of SOGATs are not well-behaved, as there is no
good notion of homomorphism between them. This is why Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: a model of a
SOGAT then is a model of the GAT the we obtain by their
translation. GATs come with well-behaved metatheory: a category of
models with initial model (syntax), (co)free models
\cite{andras,DBLP:phd/hal/Moeneclaey22}, and so on. However, the
translation is not unique: Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} define two different translations, one
based on parallel substitutions and one based on single
substitutions. If we apply their parallel translation to the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set
\end{alignedat}
\end{equation}
with only two sorts and no operations or equations, we obtain the GAT
of category with families (CwF). If we apply their single substitution
translation to the SOGAT (\ref{eq:tt}), we obtain Definition
\ref{def:tt} (also listed in Appendix \ref{app:tt}). The current paper
can be seen as taking a big enough type theory (described as a SOGAT),
and investigating how the result of its single substitution
translation relates to the result of its parallel translation (the
latter first-order theory is quite well-known). There is no general
result about the relationship of the two translations.

@@

First of all, we need a sort of terms which have to be indexed by
types because we only want well-typed terms. Both types and terms can
include variables, and to keep track of the currently available
variables, we also index them by contexts: a context is a list of
types, the length is the number of available variables, and we add new
variables at the end (that is, they are snoc-lists rather than
cons-lists). Types are also indexed by their universe level, this is a
technical requirement for avoiding Russell's paradox. The index $i$ is
an implicit argument of $\Tm$.
\begin{alignat*}{10}
& \Con && : \Set \\
& \Ty && : \Con\ra\N\ra\Set \\
& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
\end{alignat*}
Just as lists have two constructors, there are two ways to form
contexts: the empty context $\diamond$ and context extension $\ext$
which is like the snoc operation for lists. Context extension takes a
type which can have variables in the context preceeding the type.
\begin{alignat*}{10}
& \diamond && : \Con \\
& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
\end{alignat*}
We will refer to variables by their distance from the end of the
context. We would like to describe the operation providing the last
variable in a context, the zero De Bruijn index \cite{debruijn}, we denote it by $\q :
\Tm\,(\Gamma\ext A)\,A'$. It is a term in an extended context, and it
should have type $A$, but the issue is that $A : \Ty\,\Gamma$, and $A'
: \Ty\,(\Gamma\ext A)$. We need a way to weaken the type $A$ to obtain
such an $A'$. For this, we will introduce an operation $\blank[\p] :
\Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext A)\,i$ and define $A' := A[\p]$. Instead
of introducing just $\blank[\p]$, we generalise a bit and add a new
sort $\Sub$ which for now only has the single element $\p$ and the
operation $\blank[\blank]$ is called \emph{instantiation}. We
instantiate the type with the $\Sub\,\Delta\,\Gamma$ weakening.
Elements of $\Sub$ will be later called
substitutions, hence the name, but for now, we only have single end-of
context weakenings in $\Sub$. $\Delta$ is called the domain, $\Gamma$
the codomain of a $\Sub\,\Delta\,\Gamma$.
\begin{alignat*}{10}
& \Sub && : \Con\ra\Con\ra\Set \\
& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
& \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
\end{alignat*}
The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
implicitly. Still, we only have the last variable $\q$ in the context,
we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
with its context. To obtain more variables, we also allow weakening of
terms, more precisely, we introduce an instantiation operation for
terms with an overloaded notation.
\begin{alignat*}{10}
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
\end{alignat*}
Note that this is a dependent function as the type of the resulting
term has to be weakened the same way as the term itself. Now we can
define all variables counting from the end of the context (De Bruijn
indices): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
$3 := \q[\p][\p][\p]$, and so on.

@@

We add the computation and uniqueness rule for function space, where
the uniqueness rule again needs an extra equation on the codomain type
of the function to make sense. This is the type version of the
equation mentioned in the introduction (we don't have its term version).
\begin{alignat*}{10}
& \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
& \Pi\eta && : t = \lam\,(t[\p]\cdot\q)
\end{alignat*}

@@

\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
This concludes the first-order single substitution semantics of the
SOGAT (\ref{eq:tt}). For reference, we list the same rules again in
Appendix \ref{app:tt}.
\end{definition}
The theory can be extended with new type and term formers in the same
fashion: each operation has to be indexed over arbitrary contexts and come
equipped with a substitution rule. There is no need to add more
structural (substitution calculus) rules. For example, we show how to
extend Definition \ref{def:tt} with $\top$ and $\Sigma$ types.
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
  We extend Definition \ref{def:tt} with the following rules.
  \begin{alignat*}{10}
    & \top && : \Ty\,\Gamma\,0 && \top[] && : \top[\gamma] = \top \\
    & \tt && : \Tm\,\Gamma\,\top && \tt[] && : \tt[\gamma] = \tt \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong \\
    & && \rlap{$\hphantom{{}:{}} (a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : (\blank,\blank)$} \\
    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
  \end{alignat*}
  To save space, we compressed the introduction, elimination, $\beta$
  and $\eta$ rules of $\Sigma$ into one isomorphism, and did not list
  substitution laws for $\fst$ and $\snd$ as they are derivable (see
  the examples below).
\end{definition}

@@

The polymorphic identity function for types at the level 0 is defined
as
\[
\lam\,(\lam\,\q) : \Tm\,\diamond\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)\Big).
\]
Note that readability of this term essentially relies on implicit
arguments. For example, $\q$ takes $(\diamond\ext\U\,0)$ as its first,
$1$ as its second and $\Lift\,(\El\,\q)$ as its third implicit
argument, and these arguments themselves are terms written using
implicit arguments. Inrinsically typed terms are the same as
derivation trees, we illustrate this by deriving this term as follows.
\[
\infer[\text{(*)}]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
\]
Note that in step (*), $\lam$ made it sure that $\U\,0$ and
$\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
this is why we had to lift $\El\,\q$. Another subtlety is
happening when deriving the third implicit argument of $\q$. We have
to coerce $\q$ along the equality $\U[]$.
\[
\infer{\Lift\,(\El\,\q) : \Ty\,(\diamond\ext\U\,0)\,1}{\infer{\El\,\q : \Ty\,(\diamond\ext\U\,0)\,0}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,(\U\,0)}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,((\U\,0)[\p])}{} && \infer{\U[] : (\U\,0)[\p] = \U\,0}{}}}}
\]
When working informally (or in extensional type theory), we don't
write the coercion, but Agda requires it.

>>>>

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck the other equations. SSC provides a simple, minimalistic
alternative to parallel substitution calculi or B-systems for defining
type theory. SSC relates to CwF as extensional combinatory calculus
relates to lambda calculus: there are more models of the former, but
the syntaxes are equivalent. If we have some additional type formers,
we show that SSC models are actually equivalent to CwF-based models in
a weak sense. Sections 2 and 3 of this paper were formalised in Agda.
\end{abstract}

@@

The GAT presentation has the following advantages:
\begin{itemize}
\item  In type theory, typing depends on conversion:
  in an extrinsic presentation, this is expressed by a rule such as
  the following.
  \[
  \infer{\Gamma\vdash t : B}{\Gamma\vdash A \sim B && \Gamma\vdash t : A}
  \]
  When conversion $\sim$ is equality, this rule holds by the general
  properties of equality; and we also avoid lots of boilerplate
  stating that e.g.\ conversion is a congruence with respect to all
  operations (for an intrinsic presentation with explicit conversion
  relation see \cite{chapman09eatitself}).
\item Syntax and semantics are concordant: the syntax is simply the
  initial model, which always exists for any GAT and is called a
  quotient inductive-inductive type
  \cite{DBLP:journals/pacmpl/KaposiKA19}. The fact that the syntax is
  quotiented means that any function defined on it needs to respect
  the equations. For example, it is not possible to define a printing
  function which distinguishes the terms $(\lambda x.x)\,1$ and $1$,
  as these are convertible. However, it is still possible to define
  normalisation \cite{lmcs:4005} and typechecking \cite{typecheck} at
  this level of abstraction.
\item The GAT approach is also more abstract than extrinsic
  approaches, which means that we don't have to make ad-hoc
  choices. For example, we do not have to decide whether we want
  function space {\`a} la Curry or {\`a} la Church; whether we work
  with paranoid or economic typing rules \cite[Section
    5]{winterhalter:hal-01626651}; whether we have explicit or
  recursively defined substitution (instantiation of variables by
  terms, written $t[x\mapsto u]$ or $t[u/x]$).  These choices are
  forced by the notion of GAT: the domain of a lambda appears in its
  type, so we are always {\`a} la Church; algebraic theories don't
  allow recursively defined operators, so instantiation is an explicit
  operation; and so on.
\end{itemize}

@@

  \node (c2) at (5,1) {$\tilde{B}_2$};

@@

Using our notation, a B-frame contains the following data ($\diamond$ is the empty context, $\ext$ is context extension). \\

@@

Kaposi and Luksa \cite{luksa} defined a telescopic single substitution
calculus for simple type theory, it can be seen as a simply typed
version of B-systems. They show that the category of contextual models
of their calculus is equivalent to the category of contextual simply
typed CwFs with function space. Their equivalence result holds even
without the presence of function types. In the presence of function
space, the same equivalence holds for the simply typed version of our
substitution calculus, see the formalisation accompanying this paper.

@@

\paragraph{Formalisations of type theory}
Most computer formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself,setoidhell}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than plain Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. Although we don't have a general
proof, we have experimental evidence that QIITs are supported by the
cubical set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid
model \cite{kaposi-qiit-setoid}, thus Cubical Agda
\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
theory have been formalised \cite{cohtt}. Without QIITs, tricks such
as shallow embedding the syntax can be used to formalise metatheoretic
results about type theory such as gluing
\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

\paragraph{SOGATs}
We can define languages with binders as second-order theories
following higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99}
and logical frameworks
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}. Second-order
generalised algebraic theories (SOGATs
\cite{uemura,DBLP:conf/fscd/KaposiX24}) can be used to present type
theory more abstractly than GATs: SOGATs abstract over the exact
definition of the substitution calculus (whether substitutions are
single or parallel; whether we have CwF-style or Ehrhard-style
operations): contexts, substitutions, instantiations are simply
modelled by the metatheoretic function space. For example, a type
theory with $\Pi$, Coquand-universes and a universe level lifting
operation is described by the following SOGAT (where $\cong$ means
isomorphism):
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
% Adding $\top$ and $\Sigma$ types (Definition \ref{def:sigma}) is
% described as the following extension of the above SOGAT:
% \begin{equation}\label{eq:sigma}
%   \begin{alignedat}{10}
%     & \top && : \Ty\,0 \\
%     & \tt && : \Tm\,\top \\
%     & \top\eta && : t = \tt \\
%     & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
%     & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
% \end{alignedat}
% \end{equation}
Second-order models of SOGATs are not well-behaved, as there is no
good notion of homomorphism between them \cite[bottom of page 5]{DBLP:conf/fscd/KaposiX24}. This is why Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: a model of a
SOGAT then is a model of the GAT the we obtain by their
translation. GATs come with well-behaved metatheory: a category of
models with initial model (syntax), (co)free models
\cite{andras,DBLP:phd/hal/Moeneclaey22}, and so on. However, the
translation is not unique: Kaposi and Xie
\cite{DBLP:conf/fscd/KaposiX24} define two different translations, one
based on parallel substitutions and one based on single
substitutions. If we apply their parallel translation to the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set
\end{alignedat}
\end{equation}
with only two sorts and no operations or equations, we obtain the GAT
of category with families (CwF). If we apply their single substitution
translation to the SOGAT (\ref{eq:tt}), we obtain Definition
\ref{def:tt} (also listed in Appendix \ref{app:tt}). The current paper
can be seen as taking a big enough type theory (described as a SOGAT),
and investigating how the result of its single substitution
translation relates to the result of its parallel translation (the
latter first-order theory is quite well-known). There is no general
result about the relationship of the two translations.

@@

First of all, we need a sort of terms which have to be indexed by
types because we only want well-typed terms. Both types and terms can
include variables, and to keep track of the currently available
variables, we also index them by contexts: a context is a list of
types, the length is the number of available variables, and we add new
variables at the end (that is, they are snoc-lists rather than
cons-lists). Types are also indexed by their universe level, this is a
technical requirement for avoiding Russell's paradox. The index $i$ is
an implicit argument of $\Tm$.
\begin{alignat*}{10}
& \Con && : \Set \\
& \Ty && : \Con\ra\N\ra\Set \\
& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
\end{alignat*}
Just as lists have two constructors, there are two ways to form
contexts: the empty context $\diamond$ and context extension $\ext$
which is like the snoc operation for lists. Context extension takes a
type which can have variables in the context preceeding the type.
\begin{alignat*}{10}
& \diamond && : \Con \\
& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
\end{alignat*}
We will refer to variables by their distance from the end of the
context. We would like to describe the operation providing the last
variable in a context, the zero De Bruijn index \cite{debruijn}, we
denote it by $\q : \Tm\,(\Gamma\ext A)\,A'$. It is a term in an
extended context, and it should have type $A$, but the issue is that
$A : \Ty\,\Gamma$, and $A' : \Ty\,(\Gamma\ext A)$. We need a way to
weaken the type $A$ to obtain such an $A'$. For this, we will
introduce an operation $\blank[\p] : \Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext
A)\,i$ and define $A' := A[\p]$. Instead of introducing just
$\blank[\p]$, we generalise a bit and add a new sort $\Sub$ which for
now only has the single element $\p$ and the operation
$\blank[\blank]$ is called \emph{instantiation}. We instantiate the
type with the $\Sub\,\Delta\,\Gamma$ weakening.  Elements of $\Sub$
will later be called substitutions, hence the name, but for now, we
only have single end-of context weakenings in $\Sub$. The context
$\Delta$ is called the domain, $\Gamma$ the codomain of a
$\Sub\,\Delta\,\Gamma$.
\begin{alignat*}{10}
& \Sub && : \Con\ra\Con\ra\Set \\
& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
& \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
\end{alignat*}
The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
implicitly. Still, we only have the last variable $\q$ in the context,
we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
with its context. To obtain more variables, we also allow weakening of
terms, more precisely, we introduce an instantiation operation for
terms with an overloaded notation.
\begin{alignat*}{10}
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
\end{alignat*}
Note that this is a dependent function as the type of the resulting
term has to be weakened the same way as the term itself. Now we can
define all variables counting from the end of the context (De Bruijn
indices): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
$3 := \q[\p][\p][\p]$, and so on.

@@

We add the computation and uniqueness rule for function space, where
the uniqueness rule again needs an extra equation on the codomain type
of the function to make sense (we mentioned the term version of this equation in the introduction).
\begin{alignat*}{10}
& \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
& \Pi\eta && : t = \lam\,(t[\p]\cdot\q)
\end{alignat*}

@@

\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
This concludes the first-order single substitution semantics of the
SOGAT (\ref{eq:tt}). For reference, we list the same rules again in
Appendix \ref{app:tt}.
\end{definition}
The theory can be extended with new type and term formers in the same
fashion: each operation has to be indexed over arbitrary contexts and come
equipped with a substitution rule. There is no need to add more
structural (substitution calculus) rules.
\footnote{This is made formal in \cite{DBLP:conf/fscd/KaposiX24} where
these equations are enough to describe the first-order theory of any SOGAT.}
For example, we show how to
extend Definition \ref{def:tt} with $\top$ and $\Sigma$ types.
\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
  We extend Definition \ref{def:tt} with the following rules.
  \begin{alignat*}{10}
    & \top && : \Ty\,\Gamma\,0 && \top[] && : \top[\gamma] = \top \\
    & \tt && : \Tm\,\Gamma\,\top && \tt[] && : \tt[\gamma] = \tt \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong \\
    & && \rlap{$\hphantom{{}:{}} (a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : (\blank,\blank)$} \\
    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
  \end{alignat*}
  To save space, we compressed the introduction, elimination, $\beta$
  and $\eta$ rules of $\Sigma$ into one isomorphism, and did not list
  substitution laws for $\fst$ and $\snd$ as they are derivable (see
  the examples below).
\end{definition}

@@

The polymorphic identity function for types at the level 0 is defined
as
\[
\lam\,(\lam\,\q) : \Tm\,\diamond\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)\Big).
\]
Note that readability of this term essentially relies on implicit
arguments. For example, $\q$ takes $(\diamond\ext\U\,0)$ as its first,
$1$ as its second and $\Lift\,(\El\,\q)$ as its third implicit
argument, and these arguments themselves are terms written using
implicit arguments. Inrinsically typed terms are the same as
derivation trees, we illustrate this by deriving this term as follows.
\[
\infer[\text{(*)}]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
\]
Note that in step (*), $\lam$ made it sure that $\U\,0$ and
$\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
which is why we had to lift $\El\,\q$. Another subtlety is
happening when deriving the third implicit argument of $\q$, because we have
to coerce $\q$ along the equality $\U[]$:
\[
\infer{\Lift\,(\El\,\q) : \Ty\,(\diamond\ext\U\,0)\,1}{\infer{\El\,\q : \Ty\,(\diamond\ext\U\,0)\,0}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,(\U\,0)}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,((\U\,0)[\p])}{} && \infer{\U[] : (\U\,0)[\p] = \U\,0}{}}}}
\]
When working informally (or in extensional type theory), we don't
write the coercion, but Agda requires it.
